---
title: "Introduction"
author: "by Nicholas Capofari"
date: "January 22, 2017"
output: 
  html_document:
    theme: united
    toc: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      warning=FALSE, message=FALSE, comment=NA)
```

#Introduction: Wayne Winston's Mathletics in R

"Mathletics", by Wayne Winston, is a wonderful book that explains how data analytics is used in professional sports.  The data work in the book is done using Microsoft Excel.  I am an R user.  While reading the first few chapters of his book, I began thinking about how the analytic examples he uses could also serve as great introductory examples for anyone interested in learning R.

R is a freely available programming language used for statistical computing.  R is appealing because of the price, the fact that it is open source (it is constantly being improved by an engaged community of R users around the world), and R's awesome graphic capabilities.  R is quickly becoming the go-to tool for data analytics.  I have attempted to take the examples from "Mathletics" and use them to create an R tutorial.  

###Getting Started

To install R, follow this [link](https://ftp.ussg.iu.edu/CRAN/).  Once you have finished, you should also install [RStudio](https://www.rstudio.com/products/rstudio/download/).  RStudio is a an awesome intergated devlopment environment (IDE) for programming in R.

Below are some R basics.  A longer overview can be found [here](https://cran.r-project.org/doc/contrib/Torfs+Brauer-Short-R-Intro.pdf).  R's introduction manual can be found [here](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf).  I found both of these extremely helpful while coming up with these examples and explanations.

###Some R Basics

R has a robust library of open source packages.  They are easy to install.  Make sure you load the package after installation is finished.

```{r eval=FALSE} 
install.packages("Lahman") #installs the package
library(Lahman) #this loads it
```

This package provides the tables from the [Sean Lahman Baseball Database](http://www.seanlahman.com/baseball-archive/statistics/).  

Since this is an attempt to replicate work that was originally done using Microsoft Excel, a basic understanding of R's data frame structure is needed.  The simplest explanation of a data frame for an excel user is that data frames are essentially spreadsheets.  Each column represents a variable and each row contains all measured variables for the same unit.  

The rows and columns of a data frame are vectors.  This is the simplest R data structure.  A vector groups elements together in a specific order.  You can assign a vector to a variable using the *function* `c()`.

```{r}
#basic vectors
x = c(1, 2, 3)
```

Most R users use `<-` as the assignment operator.  I use `=`.  This [link](http://stackoverflow.com/questions/1741820/assignment-operators-in-r-and) can explain the difference.  When I first started programming in R, I found using `=` less confusing, so I have stuck with it.  But just so you know, [Google's R style guide](https://google.github.io/styleguide/Rguide.xml#assignment) recommends using `<-`, as does the R community.

```{r}
#more vectors
y = c("My", "first", "vector")
z <- c(TRUE, FALSE, TRUE) #same as z = c(TRUE, FALSE, TRUE)
mixed_vector = c(x[1], y[2], z[3])
mixed_vector
```

There are other ways to create vectors when the vector follows a specific pattern.

```{r}
seq(from=2, to=10, by=2)
rep(c("OH", "NO"), times=3)
c(-2:5)
```

To create a data frame, we can group vectors togther.

```{r}
Year = seq(2014, 2016)
Team = rep("New York Mets", 3)
W = c(79, 90, 87)
L = c(83, 72, 75)
mets = data.frame(Year, Team, W, L)
#if you are using RStudio, use the View() function to inspect the data frame
#View(mets)
mets
```

We can access the information in a data frame in many ways.

```{r}
#use $ to access a data frame column
mets$Year
#use logical expressions with vectors
mets$W[mets$W >= 81]
#data frame indexing: df[row, column]
mets[2, 3]
mets[ , c("W", "L")]
mets[mets$Year==2016, c(3, 4)]
mets[mets$W > mets$L, "Year"]
```

R has many built in functions.  Some functions return one element.

```{r}
max(mets$Year)
sum(mets$W)
mean(mets$W)
length(mets$Team)
```

Other functions return a vector the same length as the input.

```{r}
paste(mets$Year, mets$Team, sep="---")
#apply a fuction to every element in a vector
sapply(mets$Year, function(x){ x-2000 })
#create a new column of the fly using $
mets$Games = mets$W + mets$L
#no spaces for column names, unless column name is inside ` `
mets$W.pct = round(mets$W/mets$Games, 3)
#vectorized if 
mets$`Over 500` = ifelse(mets$W > mets$L, TRUE, FALSE)
```

We can add rows and columns to our data frames using the `rbind` and  `cbind` functions.

```{r}
next_year = c(2017, "New York Mets", 162, 0, 162, 1.000, TRUE)
rbind(mets, next_year)
cbind(mets, League=rep("NL", 3))
#our changes to the data frame were not saved 
#make sure to store any changes in a variable
mets_2 = cbind(mets, Manager=rep("Terry Collins", 3))
mets_2
```

There are many ways to import datasets into R.  `read.csv` or `readLines` can be used when we have a file we would like to work with.  Many times R packages come with ready to use datasets.

```{r}
#the cars dataset that is included with the basic R installation
data(mtcars)
#what are we wroking with?
colnames(mtcars)
nrow(mtcars)
#sneak a peek
head(mtcars) # or tail(mtcars)
#get a quick summary
summary(mtcars[ , c(1:4)])
```




